---
title: "Extended Runs 2"
author: "Philip Shea"
date: "14 October 2024"
#bibliography: references.bib
link-citations: true
format: pdf
---

```{r}
#| echo: false
#| output: false
library( parallel)
# Calculate the number of cores
num_cores <- detectCores() - 1
 source("C:/Users/phils/Documents/DSP/MD/MD2/R/runcombine.r")
```

# New Runs

The next series of runs needs to extend the runs elicited above, plus fill in the energies that are missing. The `InitMDRecord!` function was updated to zero in on the energies in a better way. Previously, the energy was set as a parameter to the random number generation, but the random generation could result in many different energies. The old code was as follows:

```         
    ss.v .= sqrt( Temp) .* randn( eltype( ss.v), ss.N)
```

The new code is as follows:

```         
    ss.v .= randn( eltype( ss.v), ss.N)
    ke = 0.5 * sum( dot.(ss.v, ss.v)) / ss.N
    ss.v .= sqrt( Temp / ke) .* ss.v
```

As can be seen, the random numbers are generated first, and then they are scaled to meet the temperature. It is still not exact, as linear and angular momentum are corrected after the kinetic energy generation, and this does change it a bit, but it ought to be a lot closer now.

Runs were made with the following.

``` julia
function runn( td::AbstractString, t::Integer, sname::AbstractString, 
               initsteps::Integer=1000, rsteps::Integer=5000, runs::Integer=3)
    s = "0123456789ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz"
    temp = t/1000
    @show temp
    tname = @sprintf "%.4d" t
    basename = sname * tname
    @show basename
    cd( td)
    io = open( basename * ".txt", "a")
    println( io, "Starting Run $basename: $(now())")
    startindx = 1
    MD2.InitSystem( basename * '.' * s[ startindx], temp=temp, io=io)
    println( io, "Starting Run $basename: $(now())")
    MD2.RunSystem( basename * '.' * s[ startindx], 
                   basename * '.' * s[ startindx += 1], initsteps, io=io)
    for i in 1:runs
        MD2.RunSystem( basename * '.' * s[ startindx], 
                   basename * '.' * s[ startindx += 1], rsteps, io=io)
        println( io, "extension finished: $(now())")
    end
    close( io)
end

function newruns(td::AbstractString, sname::AbstractString, 
                tlist::Vector{<:Integer}, initsteps::Integer, rsteps::Integer, 
                runs::Integer = 3)
    Threads.@threads for t in tlist
        runn( td, t, sname, initsteps, rsteps, runs)
    end
end

tlist = [i for i=26:38]
@time newruns( td, "S2b.", tlist, 1000, 5000, 10)
```

This last line creates a 1k initialization run, plus ten 5k runs for a total of 51k steps. The runs were created in the `Data` directory, and will be moved to the 1980 directory when they have been sufficiently processed. I expect very long runs required to achieve equilibrium. The files are processed below.

```{r}
MD2DF.S2b <- MD2DFfile( "MD2DF.S2b.RDS", "H:/Data/DSP/MD/Data", 
                       "S2b.*.scalars.csv")
```

That is a new function, created for a variety of reasons. Now, the script can be run, and `createdf` will only be called if there is a new file to be made. Here is a summary of the main parameters:

```{r}
op= par( mfrow = c(4, 2), mar = c(3, 3, 0.5, 0.5), tcl = -0.3,
mgp = c(1.7, 0.4, 0) )
plotScalars( MD2DF.S2b)
par(op)
```

Only three runs are considered valid, and we ought to be suspicious of these.

```{r}
with( MD2DF.S2b, MD2DF.S2b[Fpvalue > 0.5, 
            c("TE.mean.rs", "energy", "steps.avg")])
```

```{r}
#| label: fig-xn26
#| output: true
#| fig-cap: "Plots of the basic data in run S2b.0026."
xn26 = combineruns( "H:/Data/DSP/MD/Data", "S2b\\.0026\\.")
plotMD2( xn26, dt=0.01)
```

```{r}
#| label: fig-xn36
#| output: true
#| fig-cap: "Plots of the basic data in run S2b.0036."
xn36 = combineruns( "H:/Data/DSP/MD/Data", "S2b\\.0036\\.")
plotMD2( xn36, dt=0.01)
```

```{r}
#| label: fig-xn37
#| output: true
#| fig-cap: "Plots of the basic data in run S2b.0037."
xn37 = combineruns( "H:/Data/DSP/MD/Data", "S2b\\.0037\\.")
plotMD2( xn37, dt=0.01)
```

Those three actually look pretty good.

## Extending the rest

Ran 20 more runs of the following files: \["0027", "0028", "0029", "0030", "0031", "0032", "0033", "0034", "0035", "0038"\] (used `script5.jl` after adding some of the script functions created earlier to the code base). Note that I forgot one file, so it had to be run by itself. Below is the new standard for extending runs.

``` julia
include("C:\\Users\\phils\\Documents\\DSP\\MD\\dev\\MD2\\src\\MD2.jl")
using .MD2

origin = pwd()

td = "H:\\Data\\DSP\\MD\\Data"
cd( td)

list1 = ["0027", "0028", "0029", "0030", "0031", "0032", "0033", "0034", "0035"]

sname = "S2b."
start = 'B' # Single quotes make it a single char
@time MD2.filerun_thread( td, list1, sname, start, 5000, 20 )

@time MD2.extendrun(td, sname, "0038", 'B', 5000, 20)
```

Note that we are at "U" in the index, with only 31 more to go. The next set of extended runs will have to use larger step blocks.

```{r}
MD2DF.S2b.2 <- MD2DFfile( "MD2DF.S2b.2.RDS", "H:/Data/DSP/MD/Data", 
                       "S2b.*.scalars.csv")
```

Here are the states of the main scalars. Note that we've replaced the TE.mean vs TE.Init with the ke.rrange. This is the relative range of the kinetic energy, and is computed by

``` r
MD2DF$ke.rrange <- (MD2DF$ke.max.raw - MD2DF$ke.min.raw) / MD2DF$ke.mean.rs
```

So we can see the full range of the kinetic energy.

```{r}
op= par( mfrow = c(4, 2), mar = c(3, 3, 0.5, 0.5), tcl = -0.3,
mgp = c(1.7, 0.4, 0) )
plotScalars( MD2DF.S2b.2)
par(op)
```

We still have seven runs that are not considered in equilibrium.

```{r}
with( MD2DF.S2b.2, MD2DF.S2b.2[Fpvalue < 0.5, 
            c("TE.mean.rs", "energy", "steps.avg")])
```

Well, I tested a new feature here. It turns out that lower case letters are not unique to the file system on *Windows*. I think I knew this, but forgot. Fortunately, *Julia* made new files with, for example, `S2b.00038.a(1).csv` as the filename instead of crashing. but now I'm left with figuring out what to do with these files.

I should probably just open them all with `combineruns` and sort the total open series.  However, the step starts at 1 in each .csv file.  It was foolish to keep the entire state, but not the steps between runs. This requires a change to the MD2 program to keep track of the steps. A 64 bit integer will be enough, as even if the program were a million times faster than the current 0.1s per thousand steps, it would still take 29 years to run out of count space ( 0.1 * 2^63 / (1000000 * 1000 * 3600 * 24 * 365) = 29.5). Once this is done, future runs can be combined on step number. Until then, I will create a new version of `combineruns` that sorts by date & time using ` files <- files[ order( file.mtime( files))]`. After `MD2` is updates and a new series of runs are started, `combineruns` will need another update.

```{r}
MD2DF.S2b.3 <- MD2DFfile( "MD2DF.S2b.3.RDS", "H:/Data/DSP/MD/Data", 
                       "S2b.*.scalars.csv")
```
And here are the scalars:

```{r}
op= par( mfrow = c(4, 2), mar = c(3, 3, 0.5, 0.5), tcl = -0.3,
mgp = c(1.7, 0.4, 0) )
plotScalars( MD2DF.S2b.3)
par(op)
```

A list of the still "not valid" runs:

```{r}
with( MD2DF.S2b.3, MD2DF.S2b.3[Fpvalue < 0.5, 
            c("TE.mean.rs", "energy", "steps.avg")])
```

Only "0033" reached validity. 

```{r}
#| label: fig-xn38
#| output: true
#| fig-cap: "Plots of the basic data in run S2b.0038."
xn38 = combineruns( "H:/Data/DSP/MD/Data", "S2b\\.0038\\.")
plotMD2( tail( xn38, -1000), dt=0.01)
```

```{r}
plotMD2( tail( xn38, - 2000), dt=0.01)
```

```{r}
#| label: fig-rx38ke
#| fig-cap: "S2.0031 (@fig-xn38) kinetic energy (less the first 2,000 points) with fits to the resampled data, the black line is the fit to all of the data (red and blue circles), and the last two-thirds of truncated run S2-0031 (blue circles and line). Green lines indicate the mean (solid), and min & max (dashed) of the blue points. The red line is the regression line for the red points."
#| 
navg <-1000
skip <- 1000
(t <- procMD2DF( xn38, skip=skip, rsblk=navg))
rxn38ke <- rsmpavg( tail( xn38$ke, -skip), navg)
rxn38 <- data.frame( ke=rxn38ke, 
                    step=seq( from=navg/2, by=navg, along.with=rxn38ke))
lmrxn38 <- lm(ke ~ step, data = rxn38) # fit across entire data set
(slmrxn38 <- summary( lmrxn38))
#lenrx31 <- nrow( rx31)
#third <- floor(lenrx31 / 3) # we'll exclude the least number of points
trxn38 <- tail( rxn38, t['nr']) # procMD2DF fit
sstrxn38 <- stat.summary( trxn38$ke)
lmtrxn38 <- lm( ke ~ step, data=trxn38) # fit to tail
summary( lmtrxn38)
#ftlmxn38 <- lm( ke ~ step, data=head( rxn38, -t['nr'])) # Fit to head of data.
plot( tail( xn38$ke, -skip), type='l', panel.first=grid()) # plot raw data
points(ke ~ step, data=rxn38, col='darkred') # Plot all resample points
abline( coef=coef( lmrxn38)) # Fit to all resample points
points(ke ~ step, data=trxn38, col='blue') # Plot tail resample points
abline( coef=coef( lmtrxn38), col='blue') # Fit to the tail resample points
abline( h=sstrxn38["mean"], col='darkgreen')
abline( h=sstrxn38["min"], col="darkgreen", lty=2)
abline( h=sstrxn38["max"], col="darkgreen", lty=2)
#abline( coef=coef( ftlmxn38), col='darkred')
```