---
title: "Regression"
author: "Phil Shea"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)
```

```{r message=TRUE, warning=TRUE, include=FALSE}
load("H:\\Data\\DSP\\MD\\Data\\1980\\MD2DF.S2.Rdata")
```


# Relationship between the Estimate, Standard Error, and Regression Coeffficient

We saw a surprising relationship between the ratio of the estimate to the standard error of the estimate ($\widehat{\beta} / S_{\widehat{\beta}}$).  Here was the surprise:

```{r}
plot( (Estimate/Std.Error) ~ r.squared, data = MD2DF)

```


The graph above was generated from data which was random, although all the data had the same number (10,001) of samples.  The data was an over-sampled time series of temperature.

## Random Relationships

To examine the effect observed above, we generated a *lot* of random data.  The function `sbor()` below will generate a random number of samples (drawn from a Poisson distribution), and will generate random data from an equation $y_i = a + b x_i + \varepsilon_i$.  The function's parameters `a` & `b` set the means of the distribution of the equation parameters.  The equation parameters are drawn from normal distributions with standard deviation of one, and $x_i$ is drawn from a standard normal distribution $N(0,1)$ . The error terms are drawn from a normal distribution with mean zero and standard deviation with a default of 1/2.  

```{r}
sbor <- function(n, a, b, sde=0.5, plt=FALSE) {
   #  a & b are the means of normal distributions of sd one,
   # n is the mean of a Poisson distribution.
   nn <- 0
   while (nn == 0) nn <- rpois( 1, lambda = n)
   aa <- rnorm( 1, mean = a)
   bb <- rnorm( 1, mean = b)
   x  <- rnorm( nn) # mean zero
   y  <- aa + bb * x + rnorm( nn, mean = 0, sd = sde)
   if (plt) {
      cat("n: ", nn, " a: ", aa, " b: ", bb, "\n")
      plot( x, y)
   }
   fit  <- lm( y ~ x)
   sfit <- summary(fit)
   c( n=nn, a=aa, b=bb, coef(fit)['(Intercept)'], coef(fit)['x'],
      Sb=sfit$coefficients['x', 'Std. Error'], r=cov(x, y))
}
```

Here is an example of one such call.  The seed is set so that we can discuss the results.

```{r}
set.seed(31394)
sbor( 100, 2, 3, plt=TRUE)
pl <- recordPlot()

```

The above call generated 96 samples of data of an equation $y_i = 2.84 + 3.86 x_i + \varepsilon_i$.  The `(Intercept)` term is the estimate of $a$, and the `x` term is the estimate of the slope ($b$).  Also in the data frame is the number of degrees of freedom (`n` above, but $n - 2$ where $n$ is the number of samples) and the regression coefficient `r`. A second call would generate a different $n$, $a$, and $b$, as well as different $\varepsilon_i$.  The other thing that `sbor()` does is return the coefficients estimated, the standard error of the slope, and the regression coefficient. 

Below we can call `sbor` 10,000 times, each call generating a random number of points with a mean of 1,000, $\bar{a} = 2$ and $\bar{b} = 3$ and generate a data frame of the results.

```{r}
func <- function(i) c(trial=i, sbor( 1000, 2, 3))
# create a function of i for parLapply to call
require( parallel) # only load if needed
cl <- parallel::makeCluster( 7)
parallel::clusterExport( cl, "sbor") # export the function.
parlist = parallel::parLapply( cl=cl, 1:10000, func)
parallel::stopCluster( cl)
newdf <- as.data.frame( t( simplify2array(( parlist))))
rm(parlist) #release this memory
```

The data shows a remarkable, although not one-to-one relationship between the ratio $\widehat{\beta} / S_{\widehat{\beta}}$ and $r$:

```{r}
newdf$rs <- newdf$r^2
newdf$BoSb <- newdf$x / newdf$Sb
plot( BoSb ~ rs, data=newdf)
```

Next we look at a situation closer to the first one, where the mean slope is zero.

```{r}
rm(newdf)
func <- function(i) c(trial=i, sbor( 1000, 2, 0))
require( parallel) # only load if needed
      cl <- parallel::makeCluster( 7)
      # registerDoParallel( cl)
      parallel::clusterExport( cl, "sbor")
      parlist = parallel::parLapply( cl=cl, 1:10000, func)
      parallel::stopCluster( cl)
   newdf <- as.data.frame( t( simplify2array(( parlist))))
   rm(parlist)
newdf$rs <- newdf$r^2
newdf$BoSb <- newdf$x / newdf$Sb
plot( BoSb ~ rs, data=newdf)
pl2 <- recordPlot()
png("rndm.png", width=480, height=240)
replayPlot(pl2)
dev.off()
```

The "randomness" of the resulting graph is increased when fewer points are used, as is shown below, where only 100 points is used for each fit.


```{r}
rm(newdf)
func <- function(i) c(trial=i, sbor( 100, 2, 0))
require( parallel) # only load if needed
      cl <- parallel::makeCluster( 7)
      # registerDoParallel( cl)
      parallel::clusterExport( cl, "sbor")
      parlist = parallel::parLapply( cl=cl, 1:10000, func)
      parallel::stopCluster( cl)
   newdf <- as.data.frame( t( simplify2array(( parlist))))
   rm(parlist)
newdf$rs <- newdf$r^2
newdf$BoSb <- newdf$x / newdf$Sb
plot( BoSb ~ rs, data=newdf)
pl2 <- recordPlot()
png("rndm100.png", width=480, height=240)
replayPlot(pl2)
dev.off()
```

Here is what it looks like when $r^2$ is plotted onteh $y$ axis.

```{r}
plot( r^2 ~ b, data=newdf)
```

Below is a plot of $S_{\widehat{\beta}}$ vs. the estimates slopes (`x` in the data frame).

```{r}
plot( Sb ~ x, data=newdf)
```


Below is the calculation of the $t$ statistic for the regression coefficient $t_r = \sqrt{n - 2} \frac{r}{\sqrt{1 - r^2}}$, followed by the plot. Remember that `n` in the data frame is the number of degrees of freedom $n - 2$.

```{r}
newdf$tr <- newdf$n * newdf$r / sqrt(1 - newdf$rs)
```

Below is a plot if the $t$ statistic for the regression coefficient $t_r$:

```{r}
plot( BoSb ~ tr, data = newdf, xlim = c(-30, 30),ylim = c(-10, 10))
```

The source of the parabola is explained by the Regression Ratios document.

# Definitions

The equation for the estimate is (stolen from Wikipedia [Simple Linear Regression](https://en.wikipedia.org/w/index.php?title=Simple_linear_regression)):

$$
\widehat{\beta} = \frac {n \sum_{i=1}^n x_i y_i - 
   \sum_{i=1}^n x_i \sum_{i=1}^n y_i }
   { n\sum_{i=1}^n x_i^2-(\sum_{i=1}^n x_i)^2 }
$$

The standard error of the estimate is:

$$
s_{\widehat{\beta}} = \sqrt{\frac{\sum_{i=1}^n \widehat{\varepsilon}_i^{\,2}} 
{(n - 1) \sum_{i=1}^n x_i ^2} }
$$

with $\widehat{\varepsilon}_i = y_i - \hat{y}_i$, So the numerator can be expanded as:

$$
\sum_{i=1}^n \widehat{\varepsilon}_i^{\,2} = \sum_{i=1}^n (y_i - \hat{y}_i)^2 
= \sum_{i=1}^n (y_i^2 - 2 y_i \hat{y}_i + \hat{y}_i^2)
$$

Since $\hat{y}_i = \widehat{\alpha} + \widehat{\beta} x_i$ the above becomes


\begin{align} \label{ratio4}
    \left(\frac{\widehat{\beta}}{S_{\widehat{\beta}}} \right)^2 / \left( \frac{ S_{x y}}{S_x S_y} \right)^2 =
    \frac{(n - 2)c_1 S_y^2}{c_1 \bar{y}^2 - c_2 \bar{y} S_{x y} + c_3 S_{x y}^2 - c_1 S_y^2}
\end{align}


where $c_1 + S_x^4$, $c_2 = 2 \bar{x} S_x^2$, and $c_3 = \bar{x}^2 + S_x^2$.  For the `MD2DF` data, $x_i = i; i=1:10,000$.

```{r}
x <- 1:10000
mx <- mean(x)
Sx2 <- var(x)
Sx4 <- Sx2^2
c2 <- 2 * mx * Sx2
c3 <- mx^2 + Sx2
cat( "x mean: ", mx, " Var(x): ", Sx2, "\n")
cat( "c1: ", Sx4, " c2: ", c2, " c3: ", c3)
```

# Distribution Plots

The below from https://www.statology.org/working-with-the-student-t-distribution-in-r-dt-qt-pt-rt/

```{r}
#Create a sequence of 100 equally spaced numbers between -4 and 4
x <- seq(-4, 4, length=100)
p <- qt(0.05, 20)
p2 <- qt( .475, 20)
x <- sort( c( x, p, -p, p2, -p2))

#create a vector of values that shows the height of the probability distribution
#for each value in x, using 20 degrees of freedom
y <- dt(x = x, df = 20)

#plot x and y as a scatterplot with connected lines (type = "l") and add
#an x-axis with custom labels
plot(x, y, type = "l", lwd = 2, axes = FALSE, xlab = "", ylab = "")
axis(1, at = -3:3, labels = c("-3s", "-2s", "-1s", "mean", "1s", "2s", "3s"))
polygon( c(-4, x[x <= p], p), c( 0, y[ x <= p], 0), col='blue')
polygon( c(-p, x[x >= -p], 4), c( 0, y[ x >= -p], 0), col='blue')
l <- min( which( x >= p2))
h <- max( which( x <= -p2))
polygon( x[c( l, l:h, h)], c( 0, y[ l:h], 0), col='green')
pl <- recordPlot()
png("t20.png", width=480, height=240)
replayPlot(pl)
dev.off()
```

The tails of the F distribution are quite heavy, so the area accumulates in very thin strips. The plot below show the 50% tails.

```{r}
#Create a sequence of 100 equally spaced numbers between 0.001 and 3
df1 <- 1
df2 <- 20
x <- seq(0.001, 3, length=1000)
cat( "5%: ", qf( 0.05, df1, df2), " 95%: ", qf( .95, df1, df2))
p <- qf( 0.5, df1, df2)
p2 <- qf( .5, df1, df2)
x <- sort( c( x, p, p2))

#create a vector of values that shows the height of the probability distribution
#for each value in x, using 20 degrees of freedom
yf <- df( x, df1, df2)

#plot x and y as a scatter plot with connected lines (type = "l") and add
#an x-axis with custom labels
plot(x, yf, type = "l", lwd = 2, axes = FALSE, xlab = "", ylab = "", ylim=c(0, 3))
axis(1, at = 0:3, labels = c( "0", "1", "2", "3"))
polygon( c(0, x[x <= p], p), c( 0, yf[ x <= p], 0), col='green')
#l <- min( which( ))
h <- length(x)
polygon( c( p2, x[ x >= p2], 5),c( 0, yf[ x >= p2], 0), col='blue')
pl <- recordPlot()
png("f20.png", width=480, height=240)
replayPlot(pl)
dev.off()
```


# Replicating `lm`

```{r}
carslm <- lm(dist ~ speed, data=cars )
(scarslm <- summary(carslm))
plot(dist ~ speed, data=cars)
abline(coef( carslm))
```

```{r}
x <- cars$speed
y <- cars$dist
nr <- nrow( cars)
sumx  <- sum( x)
sumx2 <- sum( x^2)
sumy  <- sum( y)
sumy2 <- sum( y^2)
sumxy <- sum( x * y)
all.equal( mean( x), sumx / nr) # sumx == nr * mean(x)
meanx <- mean( x)
all.equal( sumx, nr * meanx) # see?
meany <- mean( y)
# show var is n-1 normalized.
all.equal( var( x), sum( (x - meanx)^2) / (nr - 1))
all.equal( var( x), (sumx2 - nr* meanx^2)  / (nr - 1)) # either way
varx <- var( x)
vary <- var( y)
# show sd(x) is simply sqrt( var(x))
all.equal( sd( x), sqrt( varx))
sdx <- sd( x)
sdy <- sd( y)
# demonstrate cov
all.equal( cov( x, y), sum( (x - meanx)*(y - meany)) / (nr - 1))
all.equal( cov( x, y), (sumxy - nr * meanx * meany) / (nr - 1))
Sxy <- cov( x, y)
all.equal( cor(x, y), Sxy / (sdx * sdy)) # correlation coef r
rxy <- cor( x, y)
all.equal( scarslm$r.squared, rxy^2) # r^2 same as lm
b <- Sxy / sdx^2
# Slope estimate the same as lm
all.equal( coef( carslm)['speed'], b,  check.attributes = FALSE)
# Names from lm will cause the above to fail.
a <- (sumy * sumx2 - sumx * sumxy)/(nr* sumx2 - sumx^2)
all.equal( meany - b * meanx, a,  check.attributes = FALSE) # simpler way to a
all.equal( coef( carslm)['(Intercept)'], a,  check.attributes = FALSE)
yest <- a + b * x
Ei <- y - yest
sumEi2 <- sum( Ei^2)
varEi <- var( Ei)
Sb <- sqrt( sumEi2/((nr - 2) * (nr - 1) * varx))
all.equal( scarslm$coefficients['speed','Std. Error'], 
           Sb, check.attributes = FALSE)
nu2 <- nr - 2
Fy <-  nu2 * (vary - varEi) / varEi
fs <-  scarslm$fstatistic
all.equal( fs[ 'value'], Fy, check.attributes = FALSE)
pv <- pf( fs['value'], fs['numdf'], fs['dendf'], lower.tail=FALSE)
c( nr=nr, sumx=sumx, sumx2=sumx2, sumy=sumy, sumy2=sumy2, sumxy=sumxy, 
   meanx=meanx, meany=meany, varx=varx, vary=vary, sdx=sdx, sdy=sdy, 
   Sxy=Sxy, sumEi2=sumEi2, rxy=rxy, rxy2=rxy^2, a=a, b=b, Sb=Sb, Fy=Fy, pv=pv)
```

The constant $\widehat{\alpha}$ is given simply by: $\hat{\bar{x}}$

$$
\widehat{\alpha} = \bar{y} - \widehat{\beta} \bar{x}
$$

```{r}
a <- meany - b * meanx
c( a=a)
ei <- y - a - b * x # this is a vector
sqrt( sum( ei^2) / ((nr - 2) * sum((x - meanx)^2) ))
scarslm$coefficients['speed','Std. Error']
all.equal( scarslm$coefficients['speed','Std. Error'], 
           sqrt( sum( ei^2) / ((nr - 2) * sum((x - meanx)^2) ) ))
Sb <- sqrt( sum( ei^2) / ((nr - 2) * (nr - 1) * varx))
all.equal( scarslm$coefficients['speed','Std. Error'], Sb)
```


